EinsteinPy: A Community Python Package for General Relativity

### 2020-05-04 paper.tex
Paper-finalise

%% Beginning of file 'sample63.tex'
%%
%% Modified 2019 June
%%
%% This is a sample manuscript marked up using the
%% AASTeX v6.3 LaTeX 2e macros.
%%
%% AASTeX is now based on Alexey Vikhlinin's emulateapj.cls 
%% (Copyright 2000-2015).  See the classfile for details.

%% AASTeX requires revtex4-1.cls (http://publish.aps.org/revtex4/) and
%% other external packages (latexsym, graphicx, amssymb, longtable, and epsf).
%% All of these external packages should already be present in the modern TeX 
%% distributions.  If not they can also be obtained at www.ctan.org.

%% The first piece of markup in an AASTeX v6.x document is the \documentclass
%% command. LaTeX will ignore any data that comes before this command. The 
%% documentclass can take an optional argument to modify the output style.
%% The command below calls the preprint style which will produce a tightly 
%% typeset, one-column, single-spaced document.  It is the default and thus
%% does not need to be explicitly stated.
%%
%%
%% using aastex version 6.3
\documentclass{aastex63}
\usepackage{amsmath}
\usepackage[utf8]{inputenc}
\usepackage{listings}  % for code in latex
\usepackage{color}
\usepackage{hyperref}
% \usepackage{caption}

%% The default is a single spaced, 10 point font, single spaced article.
%% There are 5 other style options available via an optional argument. They
%% can be invoked like this:
%%
%% \documentclass[arguments]{aastex63}
%% 
%% where the layout options are:
%%
%%  twocolumn   : two text columns, 10 point font, single spaced article.
%%                This is the most compact and represent the final published
%%                derived PDF copy of the accepted manuscript from the publisher
%%  manuscript  : one text column, 12 point font, double spaced article.
%%  preprint    : one text column, 12 point font, single spaced article.  
%%  preprint2   : two text columns, 12 point font, single spaced article.
%%  modern      : a stylish, single text column, 12 point font, article with
%% 		  wider left and right margins. This uses the Daniel
%% 		  Foreman-Mackey and David Hogg design.
%%  RNAAS       : Preferred style for Research Notes which are by design 
%%                lacking an abstract and brief. DO NOT use \begin{abstract}
%%                and \end{abstract} with this style.
%%
%% Note that you can submit to the AAS Journals in any of these 6 styles.
%%
%% There are other optional arguments one can invoke to allow other stylistic
%% actions. The available options are:
%%
%%   astrosymb    : Loads Astrosymb font and define \astrocommands. 
%%   tighten      : Makes baselineskip slightly smaller, only works with 
%%                  the twocolumn substyle.
%%   times        : uses times font instead of the default
%%   linenumbers  : turn on lineno package.
%%   trackchanges : required to see the revision mark up and print its output
%%   longauthor   : Do not use the more compressed footnote style (default) for 
%%                  the author/collaboration/affiliations. Instead print all
%%                  affiliation information after each name. Creates a much 
%%                  longer author list but may be desirable for short 
%%                  author papers.
%% twocolappendix : make 2 column appendix.
%%   anonymous    : Do not show the authors, affiliations and acknowledgments 
%%                  for dual anonymous review.
%%
%% these can be used in any combination, e.g.
%%
%% \documentclass[twocolumn,linenumbers,trackchanges]{aastex63}
%%
%% AASTeX v6.* now includes \hyperref support. While we have built in specific
%% defaults into the classfile you can manually override them with the
%% \hypersetup command. For example,
%%
%% \hypersetup{linkcolor=red,citecolor=green,filecolor=cyan,urlcolor=magenta}
%%
%% will change the color of the internal links to red, the links to the
%% bibliography to green, the file links to cyan, and the external links to
%% magenta. Additional information on \hyperref options can be found here:
%% https://www.tug.org/applications/hyperref/manual.html#x1-40003
%%
%% Note that in v6.3 "bookmarks" has been changed to "true" in hyperref
%% to improve the accessibility of the compiled pdf file.
%%
%% If you want to create your own macros, you can do so
%% using \newcommand. Your macros should appear before
%% the \begin{document} command.
%%
\newcommand{\vdag}{(v)^\dagger}
\newcommand\aastex{AAS\TeX}
\newcommand\latex{La\TeX}

%% Reintroduced the \received and \accepted commands from AASTeX v5.2
\received{June 1, 2019}
\revised{January 10, 2019}
\accepted{\today}
%% Command to document which AAS Journal the manuscript was submitted to.
%% Adds "Submitted to " the argument.
\submitjournal{ApJ}

%% For manuscript that include authors in collaborations, AASTeX v6.3
%% builds on the \collaboration command to allow greater freedom to 
%% keep the traditional author+affiliation information but only show
%% subsets. The \collaboration command now must appear AFTER the group
%% of authors in the collaboration and it takes TWO arguments. The last
%% is still the collaboration identifier. The text given in this
%% argument is what will be shown in the manuscript. The first argument
%% is the number of author above the \collaboration command to show with
%% the collaboration text. If there are authors that are not part of any
%% collaboration the \nocollaboration command is used. This command takes
%% one argument which is also the number of authors above to show. A
%% dashed line is shown to indicate no collaboration. This example manuscript
%% shows how these commands work to display specific set of authors 
%% on the front page.
%%
%% For manuscript without any need to use \collaboration the 
%% \AuthorCollaborationLimit command from v6.2 can still be used to 
%% show a subset of authors.
%
%\AuthorCollaborationLimit=2
%
%% will only show Schwarz & Muench on the front page of the manuscript
%% (assuming the \collaboration and \nocollaboration commands are
%% commented out).
%%
%% Note that all of the author will be shown in the published article.
%% This feature is meant to be used prior to acceptance to make the
%% front end of a long author article more manageable. Please do not use
%% this functionality for manuscripts with less than 20 authors. Conversely,
%% please do use this when the number of authors exceeds 40.
%%
%% Use \allauthors at the manuscript end to show the full author list.
%% This command should only be used with \AuthorCollaborationLimit is used.

%% The following command can be used to set the latex table counters.  It
%% is needed in this document because it uses a mix of latex tabular and
%% AASTeX deluxetables.  In general it should not be needed.
%\setcounter{table}{1}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%
%% The following section outlines numerous optional output that
%% can be displayed in the front matter or as running meta-data.
%%
%% If you wish, you may supply running head information, although
%% this information may be modified by the editorial offices.
\shorttitle{EinsteinPy: Python for General Relativity}
\shortauthors{The EinsteinPy Community}
%%
%% You can add a light gray and diagonal water-mark to the first page 
%% with this command:
%% \watermark{text}
%% where "text", e.g. DRAFT, is the text to appear.  If the text is 
%% long you can control the water-mark size with:
%% \setwatermarkfontsize{dimension}
%% where dimension is any recognized LaTeX dimension, e.g. pt, in, etc.
%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\graphicspath{{./}{figures/}}
%% This is the end of the preamble.  Indicate the beginning of the
%% manuscript itself with \begin{document}.

\begin{document}

\title{EinsteinPy: Python for General Relativity}

%% LaTeX will automatically break titles if they run longer than
%% one line. However, you may use \\ to force a line break if
%% you desire. In v6.3 you can include a footnote in the title.

%% A significant change from earlier AASTEX versions is in the structure for 
%% calling author and affiliations. The change was necessary to implement 
%% auto-indexing of affiliations which prior was a manual process that could 
%% easily be tedious in large author manuscripts.
%%
%% The \author command is the same as before except it now takes an optional
%% argument which is the 16 digit ORCID. The syntax is:
%% \author[xxxx-xxxx-xxxx-xxxx]{Author Name}
%%
%% This will hyperlink the author name to the author's ORCID page. Note that
%% during compilation, LaTeX will do some limited checking of the format of
%% the ID to make sure it is valid. If the "orcid-ID.png" image file is 
%% present or in the LaTeX pathway, the OrcID icon will appear next to
%% the authors name.
%%
%% Use \affiliation for affiliation information. The old \affil is now aliased
%% to \affiliation. AASTeX v6.3 will automatically index these in the header.
%% When a duplicate is found its index will be the same as its previous entry.
%%
%% Note that \altaffilmark and \altaffiltext have been removed and thus 
%% can not be used to document secondary affiliations. If they are used latex
%% will issue a specific error message and quit. Please use multiple 
%% \affiliation calls for to document more than one affiliation.
%%
%% The new \altaffiliation can be used to indicate some secondary information
%% such as fellowships. This command produces a non-numeric footnote that is
%% set away from the numeric \affiliation footnotes.  NOTE that if an
%% \altaffiliation command is used it must come BEFORE the \affiliation call,
%% right after the \author command, in order to place the footnotes in
%% the proper location.
%%
%% Use \email to set provide email addresses. Each \email will appear on its
%% own line so you can put multiple email address in one \email call. A new
%% \correspondingauthor command is available in V6.3 to identify the
%% corresponding author of the manuscript. It is the author's responsibility
%% to make sure this name is also in the author list.
%%
%% While authors can be grouped inside the same \author and \affiliation
%% commands it is better to have a single author for each. This allows for
%% one to exploit all the new benefits and should make book-keeping easier.
%%
%% If done correctly the peer review system will be able to
%% automatically put the author and affiliation information from the manuscript
%% and save the corresponding author the trouble of entering it by hand.
\input{authors.tex}

%% Note that the \and command from previous versions of AASTeX is now
%% depreciated in this version as it is no longer necessary. AASTeX 
%% automatically takes care of all commas and "and"s between authors names.

%% AASTeX 6.3 has the new \collaboration and \nocollaboration commands to
%% provide the collaboration status of a group of authors. These commands 
%% can be used either before or after the list of corresponding authors. The
%% argument for \collaboration is the collaboration identifier. Authors are
%% encouraged to surround collaboration identifiers with ()s. The 
%% \nocollaboration command takes no argument and exists to indicate that
%% the nearby authors are not part of surrounding collaborations.

%% Mark off the abstract in the ``abstract'' environment. 
\begin{abstract}

This paper presents EinsteinPy (version 0.3), a community-developed Python
package for gravitational and relativistic astrophysics. Python is a free, easy to use high level programming language which has seen a huge expansion in the number of its users and developers in recent years. Specifically, a lot of recent studies show that the use of Python in Astrophysics and in general physics has increased exponentially. Many great frameworks came as Python packages which provide a very high level of abstraction over the dirty nitty-gritty of complex algorithms and provide an easy to use interfaceand pleasing user experience. One such example is Keras - framework for deep learning which has made deep learning so easy that a person with zero programming knowledge can also train a neural network classifier. This example really demonstrates the power of abstraction which is achievable in Python. The aim of the EinsteinPy is no different and is developed keeping in mind the state of a theoretical gravitational physicist with a little or no background in computer programming and trying to work in the field of numerical relativity or trying to use simulations in their research. Currently EinsteinPy supports simulation of time-like and null geodesics and calculate trajectories in different background geometries some of which are Schwarzschild, Kerr and KerrNewmann along with coordinate inter-conversion pipeline. It has a partially developed pipeline for plotting and visualization with dependencies on libraries like plotly, matplotlib etc. One of the unique feature of EinsteinPy is a sufficiently developed symbolic tensor manipulation utilites which is a great tool in itself for teaching yourself tensor algebra which for many beginner students can be overwhelmingly tricky. Currently EinsteinPy also provide few utility functions for hypersurface embedding of Schwarzschild spacetime which further will be extended to model gravitational lensing simulation. The current version of the library is in a state that can be used by any serious student of general relativity trying to get essence of this beautiful subject but is somewhere lost in the heavy mathematical formalism of the subject. EinsteinPy provides such students to really see through the equations and visualize whats really happening.

\end{abstract}

%% Keywords should appear after the \end{abstract} command. 
%% See the online documentation for the full list of available subject
%% keywords and the rules for their use.
\keywords{gravitational physics, astrophysics --- 
simulations --- black holes --- gravitational waves}

%% From the front matter, we move on to the body of the paper.
%% Sections are demarcated by \section and \subsection, respectively.
%% Observe the use of the LaTeX \label
%% command after the \subsection to give a symbolic KEY to the
%% subsection for cross-referencing in a \ref command.
%% You can use LaTeX's \ref and \label commands to keep track of
%% cross-references to sections, equations, tables, and figures.
%% That way, if you change the order of any elements, LaTeX will
%% automatically renumber them.
%%
%% We recommend that authors also use the natbib \citep
%% and \citet commands to identify citations.  The citations are
%% tied to the reference list via symbolic KEYs. The KEY corresponds
%% to the KEY in the \bibitem in the reference list below.

\section{Introduction} \label{sec:intro}
The relativistic theory of gravity paved its way to the ground in 1916, when Prof. Albert Einstein published his paper on the general theory of relativity \citep{Einstein:1916vd}. This elegant and rigorous framework was a generalized version of gravity-free theory - special relativity, which he published earlier in 1905. Even after almost a hundred years after Einstein wrote down the equations of General Relativity, solutions of the Einstein field equations remain extremely difficult to find beyond those which exhibit significant symmetries.\\
We can study the behavior of solutions under a high degree of symmetry considerations and could even solve analytically for highly unrealistic systems. In case of problems relevant to astrophysical and gravitational physics research, it still remains a big daunting question on how to get around the problem of solving these field equations. This question is so profound that it has a separate profound field of research which goes with the name of numerical relativity which attempts to use computation science and programming to numerically obtain solutions of the equations (which would be the background geometry) for some turbulent region where most of the interesting dynamics are happening (as we can not solve for an infinitely large grid due to restrictions imposed on us by space and time complexity and computability).\\
Numerical methods have been used to solve the Einstein equation for many decades, but the past decade has seen tremendous advances and adding to that, the interest of the community grew when the field found applications in areas of radio astronomy, cosmology, signal processing, data mining. The field of numerical relativity is growing rampantly with the vast literature on algorithms, numerical methods and theoretical formulations (from basic 3+1 decomposition formulation to more sophisticated ones) with the development of robust and involved frameworks that provide a complete programming ecosystem and have proved to be essential tools for any numerical relativity researcher.\\
Reflecting the other end of the research community stands the major section of theoretical physicists which counts the majority as a novice in programming. The head challenge is the fact that the usage of these frameworks demands heavy use of high-level programming languages like C, C++, Python, etc. Though Python provides a vast room for abstractions no library existed that focused towards numerical relativity, eventually, the need of python library dedicated to general relativity became the propellant for the team and thus they made dauntless efforts to cover the space for this Python library. Since then, EinsteinPy has seen a lot of contributions from people all over the world and many ”good to go” functionalities are already provided in this and previous versions.\\

@@

The library is as an attempt to provide programming and numerical environment for numerous numerical relativity problems like geodesics plotter, gravitational lensing and ray tracing, solving and simulating relativistic hydrodynamical equations, plotting of black hole event horizons, solving Einstein’s field equations and simulating various dynamical systems like binary merger etc.\\
EinsteinPy provides an open-source and open-development core package and aims to build an ecosystem of affiliated packages that support numeric relativity functionality in the Python programming language. ”Open development” describes a process where anybody with an internet connection can suggest changes to the source code and contribute their opinion when new features, bug fixes or other code changes, governance, or any other aspect of the development process is discussed. EinsteinPy core package is now a feature-rich library of sufficiently general tools and classes that support the development of more specialized code\\
In the further coming section, we discuss with some of the features of the current as well previous version and the future plans which are yet to be implemented in the upcoming versions.Also, we describe a few code snippets to explain the usage of the library on which more details can be found on the organization website. The development of the EinsteinPy core package began as a community-driven effort to standardize core functionality for software in Python which renders a user-friendly interface for supporting numerical relativity and relativistic astrophysics research.\\
We start this paper by describing the way EinsteinPy functions followed by the main software efforts developed by the EinsteinPy itself with proper detailed documentation of the current version release. We end with a short vision for the future of EinsteinPy to its contribution in the field of computational methodology in gravitational and relativistic Astrophysics.

\section{Data Types} \label{sec:datatypes}

The EinsteinPy package provides some core data types for calculating, operating on, visualizing geodesics and black holes. There are some data types created for symbolic manipulations and calculations of relativistic tensors. The most important data types for EinsteinPy are \texttt{Body}, \texttt{Geodesic} and \texttt{BaseRelativityTensor} objects that support defining and creating new bodies, storing and calculating geodesic for bodies, and handling respectively. The main purpose of these core classes is to standardize the way calculations and visualizations are done for various kinds of heavy cosmic bodies and black holes. The classes always maintain the consistency in units of measurement wherever applicable, and create a consistent and clean API for any non-programmer to understand. The classes are designed such that it resembles the flow of physics and are  as intuitional as possible. These core classes also include a rigid architecture for data manipulation and visualizations wherever applicable. Following sections provide deep insight into the above mentioned three code data types of EinsteinPy:

\subsection{Body} \label{subsec:Body}
The Body data type helps define the central black-hole or the heavy massed object and the objects revolving around it. 
% A new body can be defined by some of the very basic constraints:

% \begin{itemize}
% \item \texttt{name}: Body Name,
% \item \texttt{mass}: Mass of the body with attached astropy units 
% \item \texttt{R}: Radius of the body with attached astropy units
% \item \texttt{differential}: Complete position and velocity vector of the body in form of a Cartesian / Spherical / BoyerLindquist data class object from einsteinpy coordinates (optional)
% \item \texttt{a}: Spin factor (a) of the body for Kerr Solutions with attached astropy units (optional)
% \item \texttt{q}: Charge of the body with attached astropy units (optional)
% \item \texttt{is\_attractor}: A boolean variable concerned with if the body is to be supposed as a central attractor. (optional)
% \item \texttt{parent}: Parent body for the body, object of the same data class (optional)
% \end{itemize}

\subsection{Geodesic} \label{subsec:Geodesic}
The Body data type helps define the central black-hole or the heavy massed object and the objects revolving around it. 
% A new body can be defined by some of the very basic constraints:

\subsection{BaseRelativityTensor} \label{subsec:BaseRelativityTensor}
This is a base class from which other tensors in \textit{symbolic} module are derived. The user can use this class to create his/her own tensor while using the various in-built functions provided by this class. The class also maintains the indicial configuration of the tensor(contravariant and covariant indices), calculates the free variables used in describing the tensor other than those which depicts the axis of space-time itself (e.g. $t, x, y, z$). The class has a function to convert the symbolic tensorial quantities into functions where actual numerical values can be substitued.

% This class along with its parent class \textit{Tensor} have the following functions/properties,
% \begin{itemize}
% \item \texttt{order} (Property) : Returns the order of the tensor. 
% \item \texttt{config} (Property) : Returns the arrangement of contravariant(upper)/covariant(lower) indices of the tensor. For example, 'ulll'.
% \item \texttt{parent\_metric} (Property) : Returns the metric tensor of the space-time associated with the tensor, if present. Else returns \texttt{None}.
% \item \texttt{tensor} (Function) : Returns the tensorial quantity contained within the class instance.
% \item \texttt{subs} (Function) : Returns a new tensor upon substitution of current variables with new variables/values.
% \item \texttt{simplify} (Function) : Returns a tensor contain simplified expressions from the current tensor.
% \item \texttt{symbols} (Function) : Returns the symbols denoting each axis of the space-time.
% \item \texttt{tensor\_lambdify} (Function) : Returns lambdified function of symbolic tensors. This means that the returned functions can accept numerical values and return numerical quantities.
% \item \texttt{lorentz\_transform} (Function) : Returns a lorentz-transformed tensor when supplied with a transformation matrix. This function also considers the contravariance and covariance of tensors during transformation.
% \end{itemize}

\section{UNIT HANDLING} \label{sec:unithandle}
All user-facing functionality provided by einsteinpy make use of astropy.units. 
All functions and objects must have their input constrained to the appropriate type of unit (e.g., length, mass or energy). Inputs can then be provided with any units that match the required type (e.g., millimeters and inches are both valid units for length) and conversions occur automatically without user intervention. The einsteinpy.constant subpackage contains a few standard constants.

\section{Development Model} \label{sec:devmodels}
The EinsteinPy Project follows an open development model, which is widely used in the scientific Python community and other computational astrophysics packages. The project is hosted on GitHub as a public repository with restricted write access to some key people. Anyone, however is welcome and encouraged to make code contributions and suggest changes using pull requests. Since the repository containing the codebase is licenced under an open and permissive MIT licence,  the project allows commercial use, modification, sublicensing, distribution and private use as long as the copyright notice and licence is redistributed. The project uses git for distributed version control and relies on CircleCI, Codefactor, GitHub Actions etc. for running the unit tests and other CI checks for every pull request.
The code quality is maintained at the highest standards with strong control on the cyclomatic complexity, code linting, imports ordering and a proper docstring wherever necessary. Every contribution is reviewed and it must meet the following requirements:

\begin{itemize}
\item \texttt The code must strictly adhere to the PEP8 Standards, on the top of it, the code must be linted with black to save the reviewer’s time identifying the changes made and reduce the diff generated because of the patch.
\item \texttt The imports in the code must be sorted using tools such as isort. 
\item \texttt The documentation follows numpy like docstrings, and a proper reStructured Text syntax is followed for writing the documentation.
\item \texttt The changelog is populated at release-time by the lead developer and the releasing developer
\end{itemize}
EinsteinPy further follows a trunk-based development model, which ensures a new branch for every minor release and new patch release made using the minor release branch. All the latest changes are kept on the master branch on GitHub and the relevant issues are closed as soon as the pull request solving it merges onto the master. A release having the changes is made at a later date. 
EinsteinPy only supports the latest bleeding-edge Python, Numpy\citep{numpy} and Astropy\citep{astropy:2018} versions.  All the stable Python versions after Python 3.5 are supported. All the code snippets written follow the guidelines of numba\citep{Lam2015NumbaAL} in order to accelerate the Python code and make the simulations faster.

\section{The EinsteinPy Core Package} \label{sec:core}
This section explains in detail, the modules available within EinsteinPy core package, their API and the purpose they serve.
\subsection{METRIC} \label{subsec:metric}
This module captures all the geometric and causal structure of specific space-times, and being able to calculate trajectories in a given space-time. 
The module derives its name from Metric Tensor, a tensorial quantity used to describe differential lengths, especially in a curved space-time. 
However, this module serves more than its usual definition and ultimately leads us to obtain geodesics of particles in a spacetime with high curvature, where notions of Newtonian physics fail.
\subsubsection{Schwarzschild Metric \citep{Schwarzschild:1916uq}} \label{subsubsec:sch}
Karl Schwarzschild pioneered the analysis of the relation between the size of black hole and its mass and this work paved the way for the first exact solution of Einstein's Field Equation under the limited case of single spherical non-rotating, non-charged body. 
The metric equation corresponding to a space-time centred around a body in the above described conditions is given by,

@@

%The same equation can be written in a tonsorial form as follows,

>>>>

%                                                                 aa.dem
% AA vers. 9.1, LaTeX class for Astronomy & Astrophysics
% demonstration file
%                                                       (c) EDP Sciences
%-----------------------------------------------------------------------
%
%\documentclass[referee]{aa} % for a referee version
%\documentclass[onecolumn]{aa} % for a paper on 1 column  
%\documentclass[longauth]{aa} % for the long lists of affiliations 
%\documentclass[letter]{aa} % for the letters 
%\documentclass[bibyear]{aa} % if the references are not structured 
%                              according to the author-year natbib style

%
\documentclass[onecolumn]{aa}  
%\documentclass[bibyear]{aa}
%\documentclass[onecolumn]{aa}  
\usepackage{graphicx}
\usepackage{listings}
\usepackage{xspace}
\usepackage{natbib}
\usepackage{url}
\usepackage[xindy, toc, hyperfirst=false, nolist, nostyles, sanitize={name=false,description=false,symbol=false}]{glossaries}
\bibpunct{(}{)}{;}{a}{}{,} % to follow the A&A style
%
\usepackage{graphicx}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{txfonts}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\usepackage[options]{hyperref}
% To add links in your PDF file, use the package "hyperref"
% with options according to your LaTeX or PDFLaTeX drivers.
%
\begin{document}

   \title{EinsteinPy: A Community Python Package for General Relativity}

   \subtitle{}

   \author{
Shreyas Bapat\inst{\ref{inst:iitm}}  % confirmed
  \and
Ritwik Saha\inst{\ref{inst:iitm}}  % confirmed
  \and
Bhavya Bhatt\inst{\ref{inst:iitm}}  % confirmed
\and
Shilpi Jain\inst{\ref{inst:iitr}}
\and
Akshita Jain\inst{\ref{inst:iitm}}
\and
Sofía Ortín Vela\inst{\ref{inst:uz}}
\and
Priyanshu Khandelwal\inst{\ref{inst:gatech}}
\and
Jyotirmaya Shivottam\inst{\ref{inst:niser}}
\and
Jialin Ma\inst{\ref{inst:gatech}}
}
   \institute{
  School of Computing and Electrical Engineering, Indian Institute of Technology Mandi, Mandi, Himachal Pradesh, India
  \label{inst:iitm}
    \and
  Department of Earth Sciences, Indian Institute of Technology Roorkee, Roorkee, Uttarakhand, India
  \label{inst:iitr}
    \and
  University of Zaragoza, Calle de Pedro Cerbuna, 12, 50009 Zaragoza, Spain
  \label{inst:uz}
    \and
  Georgia Institute of Technology, USA
  \label{inst:gatech}
    \and
   School of Physical Sciences, National Institute of Science Education and Research, Jatni, Khurdha, Odisha, India
   \label{inst:niser}
}

   \date{Received May 04, 2020; accepted May 04, 2020}

  \abstract
{This paper presents EinsteinPy (version 0.3), a community-developed Python package for gravitational and relativistic astrophysics. Python is a free, easy to use a high-level programming language which has seen a huge expansion in the number of its users and developers in recent years. Specifically, a lot of recent studies show that the use of Python in Astrophysics and general physics has increased exponentially. Many great frameworks came as Python packages which provide a very high level of abstraction over the dirty nitty-gritty of complex algorithms and provide an easy to use interface and pleasing user experience. One such example is Keras - a framework for deep learning which has made deep learning so easy that a person with zero programming knowledge can also train a neural network classifier. This example demonstrates the power of abstraction which is achievable in Python. The aim of the EinsteinPy is no different and is developed keeping in mind the state of a theoretical gravitational physicist with little or no background in computer programming and trying to work in the field of numerical relativity or trying to use simulations in their research. Currently, EinsteinPy supports simulation of time-like and null geodesics and calculates trajectories in different background geometries some of which are Schwarzschild, Kerr, and KerrNewmann along with coordinate inter-conversion pipeline. It has a partially developed pipeline for plotting and visualization with dependencies on libraries like Plotly, matplotlib, etc. One of the unique features of EinsteinPy is a sufficiently developed symbolic tensor manipulation utilities which are a great tool in itself for teaching yourself tensor algebra which for many beginner students can be overwhelmingly tricky. EinsteinPy also provides few utility functions for hypersurface embedding of Schwarzschild spacetime which further will be extended to model gravitational lensing simulation. The current version of the library is in a state that can be used by any serious student of general relativity trying to get the essence of this beautiful subject but is somewhere lost in the heavy mathematical formalism of the subject. EinsteinPy provides such students to really see through the equations and visualize what's really happening.}

   \keywords{
    gravitational physics --
    astrophysics --
    simulations --
    black holes --
    astropy
    }

   \maketitle
%
%-------------------------------------------------------------------

\section{Introduction}

   The relativistic theory of gravity paved its way to the ground in 1916, 
   when Prof. Albert Einstein published his paper on the general theory of 
   relativity \citeyear{Einstein:1916vd}. This elegant and rigorous framework 
   was a generalized version of gravity-free theory - special relativity, 
   which he published earlier in 1905. Even after almost a hundred years 
   after Einstein wrote down the equations of General Relativity, solutions 
   of the Einstein field equations remain extremely difficult to find beyond 
   those which exhibit significant symmetries.

   We can study the behaviour of solutions under a high degree of symmetry 
   considerations and could even solve analytically for highly unrealistic 
   systems. In case of problems relevant to astrophysical and gravitational 
   physics research, it remains a big daunting question on how to get around 
   the problem of solving these field equations. This question is so profound 
   that it has a separate field of research which goes with the name of numerical  
   relativity and attempts to use computation science and programming to numerically 
   obtain solutions of the equations (which would be the background geometry) 
   for some turbulent region where most of the interesting dynamics are happening 
   (as we can not solve for an infinitely large grid due to restrictions imposed 
   on us by space and time complexity and computation power).

   The field of numerical relativity is growing rampantly with the vast literature 
   on algorithms, numerical methods and theoretical formulations (from basic 3+1 
   decomposition formulation to more sophisticated ones) with the development of 
   robust and involved frameworks that provide a complete programming ecosystem 
   and have proved to be essential tools for any numerical relativity researcher.

   Reflecting the other end of the research community stands the major section 
   of theoretical physicists which counts the majority as a novice in programming. 
   The head challenge is the fact that the usage of these frameworks demands 
   heavy use of high-level programming languages like C, C++, Python, etc. 
   Though Python provides a vast room for abstractions, no library existed that  
   focused towards numerical relativity, eventually, the need of python library  
   dedicated to general relativity became the propellant for the team and thus 
   they made dauntless efforts to cover the space for this Python library. 
   Since then, EinsteinPy has seen a lot of contributions from people all over 
   the world and many ”good to go” functionalities are already provided in 
   this and previous versions.

@@

   The library is as an attempt to provide programming and numerical environment 
   for numerous numerical relativity problems like geodesics plotter, gravitational 
   lensing and ray tracing, solving and simulating relativistic hydrodynamical  
   equations, plotting of black hole event horizons, solving Einstein’s 
   field equations and simulating various dynamical systems like binary merger etc.

   EinsteinPy provides an open-source and open-development core package and 
   aims to build an ecosystem of affiliated packages that support numeric 
   relativity functionality in the Python programming language. EinsteinPy 
   core package is now a feature-rich library of sufficiently general tools 
   and classes that support the development of more specialized code.

   In the further coming section, we discuss with some of the features of the 
   current as well previous version and the plans which are yet to be implemented 
   in the upcoming versions. Also, we describe a few code snippets to explain 
   the usage of the library on which more details can be found on the 
   project website. We start this paper by describing the way EinsteinPy functions 
   followed by the main software efforts developed by the EinsteinPy itself 
   with proper detailed documentation of the current version release. We end 
   with a short vision for the future of EinsteinPy to its contribution in 
   the field of computational methodology in gravitational and 
   relativistic Astrophysics.

%--------------------------------------------------------------------
\section{Data Types}

%-------------------------------------- Two column figure (place early!)
  The EinsteinPy package provides some core data types for calculating, 
  operating, visualizing geodesics and black holes. There are some data 
  types created for symbolic manipulations and calculations of relativistic 
  tensors. The most important data types for EinsteinPy are \texttt{Body}, 
  \texttt{Geodesic} and \texttt{BaseRelativityTensor} objects that support 
  defining and creating new bodies, storing and calculating geodesic for 
  bodies, and handling respectively. The main purpose of these core classes 
  is to standardize the way calculations and visualizations are done for 
  various kinds of heavy cosmic bodies and black holes. The classes always 
  maintain the consistency in units of measurement wherever applicable, and 
  create a consistent and clean API for any non-programmer to understand. 
  The classes are designed such that it resembles the flow of physics and 
  are as intuitional as possible. These core classes also include a rigid 
  architecture for data manipulation and visualizations wherever applicable. 
  Following sections provide deep insight into the above mentioned three 
  core data types of EinsteinPy:

  \subsection{Body}
  The Body data type helps define an attractor i.e. the central black-hole 
  or the heavy mass object and later the objects under its influence. 
  The bodies can have their characteristics defined then and there thus 
  helping the user create the system easily.

  \subsection{Geodesic}
  A geodesic is the path that an-accelerating particle would follow. 
  In the plane, the geodesics are straight lines. On the sphere, the 
  geodesics are great circles. The Geodesic data type helps define the 
  geodesic of a body in the presence of an attractor or more aptly, 
  according to the system user created. These \texttt{Geodesic} objects 
  can be defined for any configuration and metric.

  \subsection{BaseRelativityTensor}
  This is a base class from which other tensors in \textit{symbolic} 
  module are derived. The user can use this class to create his/her 
  own tensor while using the various in-built functions provided by this 
  class. The class also maintains the indicial configuration of 
  the tensor(contravariant and covariant indices), calculates the 
  free variables used in describing the tensor other than those which 
  depicts the axis of space-time itself (e.g. $t, x, y, z$). The class 
  has a function to convert the symbolic tensorial quantities into 
  functions where actual numerical values can be substituted.

\section{Unit Handling}

All user-facing functionality provided by EinsteinPy make use of Astropy \citet{astropy:2013} units. All functions and objects must have their input 
constrained to the appropriate type of unit (e.g., length, mass or energy). 
Inputs can then be provided with any units that match the required type 
(e.g., millimeters and inches are both valid units for length) and conversions 
occur automatically without user intervention. The einsteinpy.constant 
subpackage contains a few standard constants.

\section{Development Model}
The EinsteinPy Project follows an open development model, which is widely used in the scientific Python community and other computational astrophysics packages. The project is hosted on GitHub as a public repository with restricted write access to some key people. However, anyone is welcomed and is encouraged to make code contributions and suggest changes using pull requests. Since the repository containing the codebase is licensed under an open and permissive MIT licence,  the project allows commercial use, modification, sub licensing, distribution and private use as long as the copyright notice and license are redistributed. The project uses git for distributed version control and relies on CircleCI, Codefactor, GitHub Actions etc. for running the unit tests and other CI checks for every pull request.
The code quality is maintained at the highest standards with a strong control on the cyclomatic complexity, code linting, imports ordering and a proper docstring wherever necessary. Every contribution is reviewed to further see if it fulfils the following requirements:

\begin{itemize}
\item \texttt The code must strictly adhere to the PEP8 Standards, on the top of it, the code must be lined with black to save the reviewer’s time identifying the changes made and reduce the diff generated because of the patch.
\item \texttt The imports in the code must be sorted using tools such as isort. 
\item \texttt The documentation follows numpy like docstrings, and a proper reStructured Text syntax is followed for writing the documentation.
\item \texttt The changelog is populated at release-time by the lead developer and the releasing developer
\end{itemize}
EinsteinPy further follows a trunk-based development model, which ensures a new branch for every minor release and new patch release made using the minor release branch. All the latest changes are kept on the master branch on GitHub and the relevant issues are closed as soon as the pull request solving it merges onto the master. A release having the changes is made at a later date. 
EinsteinPy only supports the latest bleeding-edge Python, Numpy\citep{numpy} and Astropy \citep{2019arXiv190710121V}{astropy:2018} versions.  All the stable Python versions after Python 3.5 are supported. All the code snippets written follow the guidelines of numba \citep{Lam2015NumbaAL} to accelerate the Python code and make the simulations faster.

\section{The EinsteinPy Core Package} 
This section explains in detail, the modules available within EinsteinPy core package, their API and the purpose they serve.
\subsection{Metric}
This module captures all the geometric and causal structure of specific space-times, and can calculate trajectories in a given space-time. 
The module derives its name from Metric Tensor, a tensorial quantity used to describe differential lengths, especially in curved space-time. 
However, this module serves more than its usual definition and ultimately leads us to obtain geodesics of particles in a space-time with high curvature, where notions of Newtonian physics fail.
\subsubsection{Schwarzschild Metric} 
Karl Schwarzschild pioneered the analysis of the relation between the size of black hole and its mass and this work paved the way for the first exact solution of Einstein's Field Equation under the limited case of single spherical non-rotating, non-charged body. 
The metric equation \citep{Schwarzschild:1916uq} corresponding to a space-time centred around a body in the above described conditions is given by,

@@

Schwarzschild Radius is a limit below which the degree of space-time curvature causes the particle to undergo irreversible gravitational collapse wherein the particle would need a speed greater than speed of light to escape singularity.
